---
title: "bdlnm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bdlnm}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(bdlnm)
library(dlnm)
library(INLA)
library(dplyr)
library(rlang)
library(tidyr)
library(splines)
library(ggplot2)
library(tictoc)
```

# london dataset

We will use the built-in dataset, `london` which contains daily temperature and mortality data from 1990 to 2012. Mortality counts are stratified by age group (less/more than 75 years). 

```{r}
london
```

# Set DLNM parameters

```{r}
dlnm_var <- list(
    var_prc = c(10, 75, 90),
    var_fun = "ns",
    lag_fun = "ns",
    max_lag = 21,
    lagnk = 3
)

seas <- ns(london$date, df = round(8 * length(london$date) / 365.25))

# Cross-basis parameters
argvar <- list(fun = dlnm_var$var_fun, 
               knots = quantile(london$tmean, 
                                dlnm_var$var_prc/100, na.rm = TRUE),
               Bound = range(london$tmean, na.rm = TRUE))

arglag <- list(fun = dlnm_var$lag_fun, 
               knots = logknots(dlnm_var$max_lag, nk = dlnm_var$lagnk))

# Prediction values
temp <- 
    c(quantile(london$tmean, seq(0, 0.9, by = 0.1)/100),
      quantile(london$tmean, seq(1, 99, by = 1)/100),
      quantile(london$tmean, seq(99.1, 100, by = 0.1)/100))

lags <- 0:dlnm_var$max_lag
```

# Fit model

```{r}
# Create crossbasis
cb <- crossbasis(london$tmean, lag = dlnm_var$max_lag, argvar, arglag)

# Omit NA's
attr_names <- names(attributes(cb))
attr_names <- attr_names[! attr_names %in% c("dim", "dimnames")]
attr <- attributes(cb)[attr_names]

cb <- cb[-c(1:max(dlnm_var$max_lag)),]

class(cb) <- c("crossbasis","matrix")
attributes(cb)[attr_names] <- attr

london_2 <- london[-c(1:dlnm_var$max_lag),]
seas <- seas[-c(1:dlnm_var$max_lag),]

# Fit the model (75+ years)
# config=TRUE is necessary if we want to sample from the posterior distribution
mod <- inla(mort_75plus ~ cb + factor(dow) + seas, data = london_2, family = "poisson", control.compute = list(config = TRUE))
```

# Predict & Plot

## From model

```{r}
tic()
cpred <- bcrosspred(cb, mod, at = temp)
toc()
```

Find each sample's Minimum Mortality Temperature (MMT):

```{r fig.width = 7, fig.height = 4}
mmt <- apply(cpred$allfit, 2, which.min)
mmt <- temp[mmt]
cen <- median(mmt) 

tibble(sample = 1:ncol(cpred$allfit), mmt = mmt) |> 
  ggplot(aes(x = mmt)) +
  geom_density(alpha = .3, fill = "#F8766D") +
  geom_vline(xintercept = cen, lty = 2) +
  labs(x = "Minimum Mortality Temperature (ºC)") + 
  theme_bw()
```

Center at the median of these values:

```{r}
cpred <- bcrosspred(cb, mod, at = temp, cen = cen)
```

Let's plot the overall efect using the `plot()` function:

```{r fig.width = 7, fig.height = 4}
plot(cpred, "overall", xlab = "Temperature (ºC)", ylab = "Relative Risk", col = 4, main="Overall", log = "y")
```

Or ggplot2:

```{r fig.width = 7, fig.height = 4}
median_all <- tibble(
  temp = temp, 
  rr = apply(cpred$allRRfit, 1, "median") 
)
  
cpred$allRRfit |>
  as.data.frame() |> 
  tibble::rownames_to_column("temp") |>
  pivot_longer(-temp, names_pattern = "(sample)(\\d+)", names_to = c(".value", "id")) |> 
  rename(rr = sample, sample = id) |> 
  mutate(temp = as.numeric(temp)) |> 
  ggplot(aes(x = temp, y = rr)) +
    geom_line(aes(group = sample), alpha = 0.6, color = "grey70") +
    geom_line(inherit = FALSE, data = median_all, aes(x = temp, y = rr), group = 1, lwd = 0.8) +
    geom_hline(yintercept = 1) +
    scale_y_continuous(trans = "log10") +
    labs(x = "Temperature (ºC)", y = "Relative risk") +
    theme_bw()
```

Let's plot the 3-D plot:

```{r fig.width = 7, fig.height = 6}
plot(cpred, zlab = "Relative risk", col = 4, lphi = 60, cex.axis = 0.6, xlab = "Temperature (ºC)", main = "3D graph of temperature effect")
```

Let's plot the contour plot:

```{r fig.width = 7, fig.height = 6}
plot(cpred, "contour", xlab = "Temperature (ºC)", ylab = "Lag", main="Contour plot")
```

Let's plot the lag-response effect setting the temperature equal to the MMT centering temperature:

```{r fig.width = 7, fig.height = 6}
plot(cpred , "slices", var = temp["99%"], col=3, ylab="RR", ci.arg=list(density=15,lwd=2),
     main=paste0("Association for a high temperature (", round(temp["99%"], 0), "ºC)"))
```

### Increase number of samples

We can increase the number of samplings with the `n_sample` argument:

```{r}
tic()
cpred <- bcrosspred(cb, mod, at = temp, n_sample = 10000)
toc()
```

```{r fig.width = 7, fig.height = 4}
mmt <- apply(cpred$allfit, 2, which.min)
mmt <- temp[mmt]
cen <- median(mmt) 

tibble(sample = 1:ncol(cpred$allfit), mmt = mmt) |> 
  ggplot(aes(x = mmt)) +
  geom_density(alpha = .3, fill = "#F8766D") +
  geom_vline(xintercept = cen, lty = 2) +
  labs(x = "Minimum Mortality Temperature (ºC)") + 
  theme_bw()
```

Center at the median of these values:

```{r}
cpred <- bcrosspred(cb, mod, at = temp, cen = cen, n_sample = 10000)
```

Let's plot the overall effect using the `plot()` function:

```{r fig.width = 7, fig.height = 4}
plot(cpred, "overall", xlab = "Temperature (ºC)", ylab = "Relative Risk", col = 4, main="Overall", log = "y")
```

Let's plot the 3-D plot:

```{r fig.width = 7, fig.height = 6}
plot(cpred, zlab = "Relative risk", col = 4, lphi = 60, cex.axis = 0.6, xlab = "Temperature (ºC)", main = "3D graph of temperature effect")
```

Let's plot the contour plot:

```{r fig.width = 7, fig.height = 6}
plot(cpred, "contour", xlab = "Temperature (ºC)", ylab = "Lag", main="Contour plot")
```

Let's plot the lag-response effect setting the temperature equal to the MMT centering temperature:

```{r fig.width = 7, fig.height = 6}
plot(cpred , "slices", var = temp["99%"], col=3, ylab="RR", ci.arg=list(density=15,lwd=2),
     main=paste0("Association for a high temperature (", round(temp["99%"], 0), "ºC)"))
```

## Supply coefficients

Alternatively, we can predict from the supplied coefficients from the posterior distribution. 

First, we have to extract the coefficients:

```{r}
names_sel <- grep("^v", rownames(mod$summary.fixed), value = TRUE)
list_sel <- lapply(1:length(names_sel), function(x) 1)
names(list_sel) <- names_sel

tic()
coef <- inla.posterior.sample(n = 100, mod, selection = list_sel)
toc()
```

Then, we perform the prediction:

```{r}
tic()
cpred <- bcrosspred(cb, coef = coef, model.link = "log", at = temp)
toc()
```

Center the results at the median MMT:

```{r}
mmt <- apply(cpred$allfit, 2, which.min)
mmt <- temp[mmt]
cen <- median(mmt)
cpred <- bcrosspred(cb, coef = coef, model.link = "log", at = temp, cen = cen)
```

Let's plot the overall effect using the `plot()` function:

```{r fig.width = 7, fig.height = 4}
plot(cpred, "overall", xlab = "Temperature (ºC)", ylab = "Relative Risk", col = 4, main="Overall", log = "y")
```

Let's plot the 3-D plot:

```{r fig.width = 7, fig.height = 6}
plot(cpred, zlab = "Relative risk", col = 4, lphi = 60, cex.axis = 0.6, xlab = "Temperature (ºC)", main = "3D graph of temperature effect")
```

Let's plot the contour plot:

```{r fig.width = 7, fig.height = 6}
plot(cpred, "contour", xlab = "Temperature (ºC)", ylab = "Lag", main="Contour plot")
```

Let's plot the lag-response effect setting the temperature equal to the MMT centering temperature:

```{r fig.width = 7, fig.height = 6}
plot(cpred , "slices", var = temp["99%"], col=3, ylab="RR", ci.arg=list(density=15,lwd=2),
     main=paste0("Association for a high temperature (", round(temp["99%"], 0), "ºC)"))
```
