---
title: "bdlnm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bdlnm}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(bdlnm)
library(dlnm)
library(INLA)
library(dplyr)
library(rlang)
library(tidyr)
library(splines)
library(ggplot2)
library(tictoc)
library(stringr)
```

# london dataset

We will use the built-in dataset, `london` which contains daily temperature and mortality data from 1990 to 2012. Mortality counts are stratified by age group (less/more than 75 years). 

```{r}
london
```

# Set DLNM parameters

```{r}
dlnm_var <- list(
    var_prc = c(10, 75, 90),
    var_fun = "ns",
    lag_fun = "ns",
    max_lag = 21,
    lagnk = 3
)

seas <- ns(london$date, df = round(8 * length(london$date) / 365.25))

# Cross-basis parameters
argvar <- list(fun = dlnm_var$var_fun, 
               knots = quantile(london$tmean, 
                                dlnm_var$var_prc/100, na.rm = TRUE),
               Bound = range(london$tmean, na.rm = TRUE))

arglag <- list(fun = dlnm_var$lag_fun, 
               knots = logknots(dlnm_var$max_lag, nk = dlnm_var$lagnk))

# Prediction values
temp <- 
    c(quantile(london$tmean, seq(0, 0.9, by = 0.1)/100),
      quantile(london$tmean, seq(1, 99, by = 1)/100),
      quantile(london$tmean, seq(99.1, 100, by = 0.1)/100))

lags <- 0:dlnm_var$max_lag
```

# Fit model

```{r}
# Create crossbasis
cb <- crossbasis(london$tmean, lag = dlnm_var$max_lag, argvar, arglag)

# Omit NA's
attr_names <- names(attributes(cb))
attr_names <- attr_names[! attr_names %in% c("dim", "dimnames")]
attr <- attributes(cb)[attr_names]

cb <- cb[-c(1:max(dlnm_var$max_lag)),]

class(cb) <- c("crossbasis","matrix")
attributes(cb)[attr_names] <- attr

london_2 <- london[-c(1:dlnm_var$max_lag),]
seas <- seas[-c(1:dlnm_var$max_lag),]

# Fit the model (75+ years)
# config=TRUE is necessary if we want to sample from the posterior distribution
mod <- inla(mort_75plus ~ cb + factor(dow) + seas, data = london_2, family = "poisson", control.compute = list(config = TRUE))
```

# Minimum mortality temperature

Let's get the minimum mortality temperature (MMT):

```{r}
mmt <- minimum_risk(cb, mod, at = temp, seed = 5243)

str(mmt)

mmt$res |> 
  head()

mmt$res.summary
```

We can plot the posterior distribution of the MMT estimates:

```{r fig.width = 7}
tibble(sample = seq_along(mmt$res), mmt = mmt$res) |> 
  ggplot(aes(x = mmt)) +
  geom_density(alpha = .3, fill = "#F8766D") +
  geom_vline(xintercept = mmt$res.summary[["0.5quant"]], lty = 2) +
  labs(x = "Minimum Mortality Temperature (ºC)") + 
  theme_bw()
```

We will center the risk at the median of these values:

```{r}
cen <- mmt$res.summary[["0.5quant"]] 
cen
```

# Risk

Predict the risk at the quantile temperature values:

```{r}
tic()
cpred <- bcrosspred(cb, mod, at = temp)
toc()
```

We can repeat it centering the risk at the MMT median values obtained before:

```{r}
tic()
cpred <- bcrosspred(cb, mod, at = temp, cen = cen)
toc()
```

We can set a seed for reproducibility:

```{r}
tic()
cpred <- bcrosspred(cb, mod, at = temp, cen = cen, seed = 5243)
toc()
```

Alternatively, we can predict from the supplied coefficients from the posterior distribution. 

First, we have to extract the coefficients:

```{r}
names_sel <- grep("^v", rownames(mod$summary.fixed), value = TRUE)
list_sel <- lapply(1:length(names_sel), function(x) 1)
names(list_sel) <- names_sel

tic()
coef <- inla.posterior.sample(n = 1000, mod, selection = list_sel, seed = 5243)
toc()
```

Then, we perform the prediction:

```{r}
tic()
cpred <- bcrosspred(cb, coef = coef, model.link = "log", at = temp, cen = cen, seed = 5243)
toc()
```

Let's see the `str` of the generated prediction:

```{r}
cpred <- bcrosspred(cb, mod, at = temp, cen = cen)

str(cpred)
```

In this object we store each of the generated samples from the posterior distribution and its summaries. 

For example, we have the estimated crossbasis coefficients:

```{r}
cpred$coefficients |> 
  head(c(5, 5))
```

We have also stored all the RR effects for each temperature and lag in each of the samples:

```{r}
cpred$matRRfit |> 
  head(c(5, 5))
```

And the overall effects for each temperature (adding up all the lags):

```{r}
cpred$allRRfit |> 
  head(c(5, 5))
```

In addition, the summary of these effects is also available:

```{r}
cpred$matRRfit.summary |> 
  head(5)

cpred$allRRfit.summary |> 
  head(5)
```

# Plot

Center at the median of these values:

```{r}
cpred <- bcrosspred(cb, mod, at = temp, cen = cen, seed = 5243)
```

Let's plot the overall effect using the `plot()` function:

```{r fig.width = 7, fig.height = 4}
plot(cpred, "overall", xlab = "Temperature (ºC)", ylab = "Relative Risk", col = 4, main="Overall", log = "y")
```

We can plot the 99% credible interval:

```{r fig.width = 7, fig.height = 4}
plot(cpred, "overall", xlab = "Temperature (ºC)", ylab = "Relative Risk", col = 4, main="Overall", log = "y", ci.level = 0.99)
```

Or we can plot all the lines from each of the posterior samples:

```{r fig.width = 7, fig.height = 4}
plot(cpred, "overall", xlab = "Temperature (ºC)", ylab = "Relative Risk", col = 4, main="Overall", log = "y", ci = "sampling")
```

Let's plot the 3-D plot:

```{r fig.width = 7, fig.height = 6}
plot(cpred, zlab = "Relative risk", col = 4, lphi = 60, cex.axis = 0.6, xlab = "Temperature (ºC)", main = "3D graph of temperature effect")
```

Let's plot the contour plot:

```{r fig.width = 7, fig.height = 6}
plot(cpred, "contour", xlab = "Temperature (ºC)", ylab = "Lag", main="Contour plot")
```

Let's plot the lag-response effect setting the temperature equal to the MMT centering temperature:

```{r fig.width = 7, fig.height = 6}
plot(cpred , "slices", var = temp["99%"], col=3, ylab="RR",
     main=paste0("Association for a high temperature (", round(temp["99%"], 0), "ºC)"))
```

Or the exposure-response at lag 0:

```{r fig.width = 7, fig.height = 6}
plot(cpred , "slices", lag = 0, col=4, ylab="RR",
     main=paste0("Association at Lag 0"))
```

# Attributable number

Calculate the fraction and number of deaths attributable to temperature in all the period. We can calculate it with the backwards algorithm:

```{r}
tic()
attributable_risk(london$tmean, cb, london$mort_75plus, mod, cen = cen, n_sample = 100, dir = "back")
toc()
```

Or with the forward one:

```{r}
tic()
attributable_risk(london$tmean, cb, london$mort_75plus, mod, cen = cen, n_sample = 100, dir = "forw")
toc()
```

We can also calculate the attributable number for each value of the temperature series:

```{r}
tic()
ar <- attributable_risk(london$tmean, cb, london$mort_75plus, mod, cen = cen, n_sample = 100, dir = "back", tot = FALSE)
toc()

ar$an |> 
  head()
```
