---
title: "bdlnm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bdlnm}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(bdlnm)
library(dlnm)
library(INLA)
library(splines)
```

# london dataset

We will use the built-in dataset, `london` which contains daily temperature and mortality data from 2000 to 2012. Using Distributed Non-Linear Models (DLNM) we will estimate the effect of the temperature exposure to mortality in the population with more than 75 years old, taking 21 days of lags.

```{r}
london
```

# Set DLNM parameters

We will start by setting the basic parameters for the configuration of the crossbasis:

```{r}
# Exposure-response and lag-response spline parameters
dlnm_var <- list(
    var_prc = c(10, 75, 90),
    var_fun = "ns",
    lag_fun = "ns",
    max_lag = 21,
    lagnk = 3
)


# Cross-basis parameters
argvar <- list(fun = dlnm_var$var_fun, 
               knots = quantile(london$tmean, 
                                dlnm_var$var_prc/100, na.rm = TRUE),
               Bound = range(london$tmean, na.rm = TRUE))

arglag <- list(fun = dlnm_var$lag_fun, 
               knots = logknots(dlnm_var$max_lag, nk = dlnm_var$lagnk))
```

With the defined parameters, we can create now the crossbasis using the {dlnm} package:

```{r}
# Create crossbasis
cb <- crossbasis(london$tmean, lag = dlnm_var$max_lag, argvar, arglag)
```

First, let's define the seasonality of the analysed mortality time series and the temperature exposure values for which we want to make predictions:

```{r}
# Seasonality of mortality time series
seas <- ns(london$date, df = round(8 * length(london$date) / 365.25))

# Prediction values (equidistant points)
temp <- round(seq(min(london$tmean), max(london$tmean), by = 0.1), 1)
```

# Fit the model and posterior samples

Fit the bayesian DLNM model with the function `bdlnm()`. We will use a Poisson model:

```{r}
# tictoc::tic()
mod <- bdlnm(mort_75plus ~ cb + factor(dow) + seas, basis = cb, data = london, family = "poisson")
# tictoc::toc()
```

We can set a seed for reproducibility:

```{r}
# tictoc::tic()
mod <- bdlnm(mort_75plus ~ cb + factor(dow) + seas, basis = cb, data = london, family = "poisson", sample.arg = list(seed = 5243))
# tictoc::toc()
```

# Minimum mortality temperature

Let's get the minimum mortality temperature (MMT), defined as the value at which the overall mortality risk is minimized:

```{r}
# tictoc::tic()
mmt <- minimum_effect(mod, cb, at = temp)
# tictoc::toc()

str(mmt)

mmt$min |> 
  head()

mmt$min.summary
```

We can plot the posterior distribution of the posterior distribution sample of the MMT estimates:

```{r fig.width = 7}
plot(mmt, xlab = "Temperature (ºC)", main = paste0("MMT (Median = ", round(mmt$min.summary[["0.5quant"]], 1), "ºC)"))
```

To make the predictions, we will center the risk at the median of these values:

```{r}
cen <- mmt$min.summary[["0.5quant"]] 
cen
```

# Predict risk

Predict the risk at the quantile temperature values:

```{r}
# tictoc::tic()
cpred <- bcrosspred(mod, cb, at = temp)
# tictoc::toc()
```

We can repeat it centering the risk at the MMT median values obtained before:

```{r}
# tictoc::tic()
cpred <- bcrosspred(mod, cb, at = temp, cen = cen)
# tictoc::toc()
```

Let's see the `str` of the generated prediction:

```{r}
str(cpred)
```

In this object we store each of the generated samples from the posterior distribution and its summaries. 

For example, we have the estimated crossbasis coefficients:

```{r}
cpred$coefficients |> 
  head(c(5, 5))
```

We have also stored all the RR effects in an array for each temperature and lag in each of the samples. For example, for the first sample:

```{r}
cpred$matRRfit[, , "sample1"] |> 
  head()
```

And the overall effects for each temperature (adding up all the lags):

```{r}
cpred$allRRfit |> 
  head(c(5, 5))
```

In addition, the summary of these effects is also available:

```{r}
cpred$matRRfit.summary |> 
  head(5)

cpred$allRRfit.summary |> 
  head(5)
```

# Plot

Let's plot the overall effect for each temperature exposure using the `plot()` function:

```{r fig.width = 7, fig.height = 4}
plot(cpred, "overall", xlab = "Temperature (ºC)", ylab = "Relative Risk", col = 4, main="Overall", log = "y")
```

We can plot the 99% credible interval, instead of the 95% used by default:

```{r fig.width = 7, fig.height = 4}
plot(cpred, "overall", xlab = "Temperature (ºC)", ylab = "Relative Risk", col = 4, main="Overall", log = "y", ci.level = 0.99)
```

We can also plot all the lines from each of the posterior samples:

```{r fig.width = 7, fig.height = 4}
plot(cpred, "overall", xlab = "Temperature (ºC)", ylab = "Relative Risk", col = 4, main="Overall", log = "y", ci = "sampling")
```

Let's plot the 3-D plot, visualizing both the exposure-response and the lag-response relationships:

```{r fig.width = 7, fig.height = 6}
plot(cpred, "3d", zlab = "Relative risk", col = 4, lphi = 60, cex.axis = 0.6, xlab = "Temperature (ºC)", main = "3D graph of temperature effect")
```

Let's plot the contour plot, which is a projection in 2D of the previous relationship:

```{r fig.width = 7, fig.height = 6}
plot(cpred, "contour", xlab = "Temperature (ºC)", ylab = "Lag", main="Contour plot")
```

Let's plot the lag-response effect for a high temperature (99th percentile):

```{r fig.width = 7, fig.height = 6}
htemp <- 23
plot(cpred , "slices", var = htemp, col=3, ylab="RR",
     main=paste0("Association for a high temperature (", htemp, "ºC)"))
```

Or the exposure-response at lag 0:

```{r fig.width = 7, fig.height = 6}
plot(cpred , "slices", lag = 0, col=4, ylab="RR",
     main=paste0("Association at Lag 0"))
```
