---
title: "Attributable measures"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bdlnm)
library(dlnm)
library(dplyr)
library(tidyr)
library(stats)
library(splines)
library(utils)
library(ggplot2)
```

This article is a short tutorial to calculate attributable measures using the `attributable()` function. We will use the built-in `london` dataset. 

First, we have to prepare the data, build the DLNM crossbasis, fit the B-DLNM model and obtain the minimum mortality temperature value to center the risks in order to calculate properly attributable measures. 

```{r warning=FALSE}
# Exposure-response and lag-response spline parameters
dlnm_var <- list(
  var_prc = c(10, 75, 90),
  var_fun = "ns",
  lag_fun = "ns",
  max_lag = 21,
  lagnk = 3
)

# Cross-basis parameters
argvar <- list(fun = dlnm_var$var_fun,
               knots = quantile(london$tmean,
                                dlnm_var$var_prc/100, na.rm = TRUE),
               Bound = range(london$tmean, na.rm = TRUE))

arglag <- list(fun = dlnm_var$lag_fun,
               knots = logknots(dlnm_var$max_lag, nk = dlnm_var$lagnk))

# Create crossbasis
cb <- crossbasis(london$tmean, lag = dlnm_var$max_lag, argvar, arglag)

# Seasonality of mortality time series
seas <- ns(london$date, df = round(8 * length(london$date) / 365.25))

# Prediction values (equidistant points)
temp <- round(seq(min(london$tmean), max(london$tmean), by = 0.1), 1)

# Model

mod <- bdlnm(mort_75plus ~ cb + factor(dow) + seas,
           basis = cb,
           data = london,
           family = "poisson",
           sample.arg = list(seed = 432))

# Predict
cpred <- bcrosspred(mod, cb, at = temp)

# compute centering (MMT) using optimal_exposure
mmt <- optimal_exposure(mod, cb, at = temp)
cen <- mmt$summary[["0.5quant"]]
cen
```

The MMT value in which values will be centered is `r cen`ºC. Now, we can calculate attributable numbers and fractions. 

We can calculate them using the backwards algorithm, which will give the current burden due to the set of exposure events experienced in the past:

```{r}
tictoc::tic()
attr_back <- attributable(mod, cb, london, name_date = "date", name_exposure = "tmean", name_cases = "mort_75plus", cen = cen, dir = "back")
tictoc::toc()
```
The output will be a list containing the posterior samples of all attributable fractions and numbers per day and for the entire period, accompanied by their summaries:

```{r}
str(attr_back)
```

Let's plot the time series of attributable fractions and numbers per day:

```{r fig.width = 10}
london |> 
  mutate(
    "Attributable Fractions" = attr_back$af.summary[,"0.5quant"],
    "Attributable Numbers" = attr_back$an.summary[,"0.5quant"]
  ) |> 
  pivot_longer(c("Attributable Fractions", "Attributable Numbers"), names_to = "name", values_to = "val") |> 
  ggplot(aes(x = date, y = val, color = name)) +
  facet_wrap(~name, scales = "free_y") +
  geom_line() +
  scale_x_date(date_breaks = "2 years") +
  labs(x = "Date", y = "") +
  guides(color = "none") +
  theme_bw()
```

The total attributable fraction and number are:

```{r}
options(scipen=999)

rbind(
  "Attributable fraction" = attr_back$aftotal.summary,
  "Attributable number" = attr_back$antotal.summary
) |> 
  knitr::kable(digits = 2)
```


Alternatively, we can use the forward algorithm which will give the future burden due to a given exposure event:

```{r}
tictoc::tic()
attr_forw <- attributable(mod, cb, london, name_date = "date", name_exposure = "tmean", name_cases = "mort_75plus", cen = cen, dir = "forw")
tictoc::toc()
```
Let's plot the time series of attributable fractions and numbers per day, with the forward algorithm:

```{r fig.width = 10}
london |> 
  mutate(
    "Attributable Fractions" = attr_forw$af.summary[,"0.5quant"],
    "Attributable Numbers" = attr_forw$an.summary[,"0.5quant"]
  ) |> 
  pivot_longer(c("Attributable Fractions", "Attributable Numbers"), names_to = "name", values_to = "val") |> 
  ggplot(aes(x = date, y = val, color = name)) +
  facet_wrap(~name, scales = "free_y") +
  geom_line() +
  scale_x_date(date_breaks = "2 years") +
  labs(x = "Date", y = "") +
  guides(color = "none") +
  theme_bw()
```

The total attributable fraction and number, with the forward algorithm, are:

```{r}
rbind(
  "Attributable fraction" = attr_forw$aftotal.summary,
  "Attributable number" = attr_forw$antotal.summary
) |> 
  knitr::kable(digits = 2)
```

Finally, we can also filter for some specific time points. For example, imagine we can calculate attributable measures only for summers:

```{r}
slondon <- london |> 
  mutate(
    summer = case_when(
      lubridate::month(date) >= 6 & lubridate::month(date) <= 8 ~ 1,
      .default = 0
    )
  )

# tictoc::tic()
attr_summer <- attributable(mod, cb, slondon, name_date = "date", name_exposure = "tmean", name_cases = "mort_75plus", name_filter = "summer", cen = cen, dir = "back")
# tictoc::toc()
```

Let's plot the time series of attributable fractions and numbers per day, only for summers:

```{r fig.width = 10}
slondon |> 
  filter(summer == 1) |> 
  mutate(
    "Attributable Fractions" = attr_summer$af.summary[,"0.5quant"],
    "Attributable Numbers" = attr_summer$an.summary[,"0.5quant"]
  ) |> 
  pivot_longer(c("Attributable Fractions", "Attributable Numbers"), names_to = "name", values_to = "val") |> 
  ggplot(aes(x = date, y = val, color = name, group = year)) +
  facet_wrap(~name, scales = "free_y") +
  geom_line() +
  scale_x_date(date_breaks = "2 years") +
  labs(x = "Date", y = "") +
  guides(color = "none") +
  theme_bw()
```

The total attributable fraction and number, only for summers, are:

```{r}
rbind(
  "Attributable fraction" = attr_summer$aftotal.summary,
  "Attributable number" = attr_summer$antotal.summary
) |> 
  knitr::kable(digits = 2)
```

