---
title: "attributable"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bdlnm)
library(dlnm)
library(dplyr)
library(stats)
library(splines)
```

This article is a short tutorial to calculate attributable measures using the `attributable()` function. We will use the built-in `london` dataset, but this time filtered for only the year `2012` to reduce the computational time of calculations. 

```{r}
# filter london dataset to run faster
slondon <- london |>
  filter(year >= 2012)

slondon
```

First, we have to prepare the data, build the DLNM crossbasis, fit the B-DLNM model and obtain the minimum mortality temperature value to center the risks in order to calculate properly attributable measures. 

```{r warning=FALSE}
# Exposure-response and lag-response spline parameters
dlnm_var <- list(
  var_prc = c(10, 75, 90),
  var_fun = "ns",
  lag_fun = "ns",
  max_lag = 21,
  lagnk = 3
)

# Cross-basis parameters
argvar <- list(fun = dlnm_var$var_fun,
               knots = quantile(slondon$tmean,
                                dlnm_var$var_prc/100, na.rm = TRUE),
               Bound = range(slondon$tmean, na.rm = TRUE))

arglag <- list(fun = dlnm_var$lag_fun,
               knots = logknots(dlnm_var$max_lag, nk = dlnm_var$lagnk))

# Create crossbasis
cb <- crossbasis(slondon$tmean, lag = dlnm_var$max_lag, argvar, arglag)

# Seasonality of mortality time series
seas <- ns(slondon$date, df = round(8 * length(slondon$date) / 365.25))

# Prediction values (equidistant points)
temp <- round(seq(min(slondon$tmean), max(slondon$tmean), by = 0.1), 1)

# Model

mod <- bdlnm(mort_75plus ~ cb + factor(dow) + seas,
           basis = cb,
           data = slondon,
           family = "poisson",
           sample.arg = list(seed = 432))

# Predict
cpred <- bcrosspred(mod, cb, at = temp)

# compute centering (MMT) using minimum_effect
mmt <- minimum_effect(mod, cb, at = temp)
cen <- mmt$min.summary[["0.5quant"]]
cen
```

The MMT value in which values will be centered is `cen`ºC. Now, we can calculate attributable number and fractions.

We can calculate it with the backwards algorithm:

```{r}
# tictoc::tic()
attr <- attributable(mod, cb, slondon, name_date = "date", name_exposure = "tmean", name_cases = "mort_75plus", cen = cen, dir = "back")
# tictoc::toc()

str(ar)
```

Or with the forward one:

```{r}
# tictoc::tic()
attr <- attributable(mod, cb, slondon, name_date = "date", name_exposure = "tmean", name_cases = "mort_75plus", cen = cen, dir = "forw")
# tictoc::toc()

str(attr)
```

We can also calculate the attributable number corresponding to each day:

```{r}
# tictoc::tic()
attr <- attributable(mod, cb, slondon, name_date = "date", name_exposure = "tmean", name_cases = "mort_75plus", cen = cen, dir = "back", tot = FALSE)
# tictoc::toc()

str(attr)
```
